Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-07-16 articles/formalism/sections/discussion.tex
minor update to references, revision to discussion surround the radial kernel relation to zaldariagga kernel

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial part of the operator kernel are roughly self-similar in the sense that the radial kernels evaluated with some band limit is related to another radial kernel evaluated with a different band limit by a simple rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ which is simply the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180, 180 \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). \revisit{An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations, and some such averaging takes place due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.}

>>>>

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial part of the operator kernel are roughly self-similar in the sense that the radial kernels evaluated with some band limit is related to another radial kernel evaluated with a different band limit by a simple rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ which is simply the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). \revisit{An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations, and some such averaging takes place due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.}

### 2018-07-16 articles/formalism/sections/kernel_discussion.tex
minor update to references, revision to discussion surround the radial kernel relation to zaldariagga kernel

\revisit{The radial function ${}_\mm f(\beta)$ is oscillatory, regardless of the maximum $l_{\rm max}$. The envelope of the oscillation decays with angular separation at intermediate $\beta$ but increases as $\beta \rightarrow \pi$.  We find that the positive envelope of ${}_{\mm}f$ scales as $\beta^{-1.55}$ at intermediate $\beta$.  However, when we average over several oscillations with a moving window, we find that the average behavior as intermediate $\beta$ scales as $\beta^{-2}$, visible in \fig{fig:fbeta_avg}.  (We have narrowed the window as the band limit increases so that the number of oscillations averaged remains the same.)  This agrees with the hypothesis for the flat sky case, for which e.g. Zaldarriaga (2001) \cite{Zaldarriaga2001a} argued that form of the radial function has to be ${}_\mm f(\beta>0)\simeq\beta^{-2}$ to ensure that the Fourier modes of Stokes $Q/U$ relate to those of $E/B$ merely by rotations (with no scale dependence). In addition, we find that ${}_{\md}f$ under a moving average shows a similar scaling of $\beta^{-2}$, as in \fig{fig:dfbeta_avg}.}

\revisit{In essence the radial function resulting from taking a sliding average can be thought of as transitioning from a higher band limit radial function to a radial function with an effectively lower band limit, set by the width of the sliding window.}
\comment{K: I don't follow this sentence.} 
\revisit{Applying a moving average on radial functions with increasing band limit, we see that a larger fraction of angular domain matches the scaling of $\beta^{-2}$.   Extrapolating this trend to the case of a very high band limit we may expect that this behavior will approach $\beta^{-2}$ across the domain $\beta\in[\epsilon,\pi-\epsilon]$ for $\epsilon \ll 1$.}

>>>>

\revisit{The radial function ${}_\mm f(\beta)$ is oscillatory, regardless of the maximum $l_{\rm max}$. The envelope of the oscillation decays with angular separation at intermediate $\beta$ but increases as $\beta \rightarrow \pi$.  We find that the positive envelope of ${}_{\mm}f$ scales as $\beta^{-1.6}$ at intermediate $\beta$.  However, when we average over several oscillations with a moving window, we find that the average behavior at intermediate $\beta$ scales as $\beta^{-2}$, visible in \fig{fig:fbeta_avg}  (Note: We have narrowed the window as the band limit increases so that the number of oscillations averaged remains roughly the same.).  This agrees with the hypothesis for the flat sky case, for which e.g. Zaldarriaga (2001) \cite{Zaldarriaga2001a} argued that form of the radial function has to be ${}_\mm f(\beta>0)\simeq\beta^{-2}$ to ensure that the Fourier modes of Stokes $Q/U$ relate to those of $E/B$ merely by rotations (with no scale dependence). In addition, we find that ${}_{\md}f$ under a moving average shows a similar scaling of $\beta^{-2}$, as seen in \fig{fig:dfbeta_avg}.} \revisit{Applying a moving average on radial functions with increasing band limit, we see that a larger fraction of angular domain matches the scaling of $\beta^{-2}$.   Extrapolating this trend to the case of a very high band limit we may expect that this behavior will approach $\beta^{-2}$ across the domain $\beta\in[\epsilon,\pi-\epsilon]$ for $\epsilon \ll 1$.} \revisit{In essence the moving average can be understood as the averaging resulting from discrete sampling of a continuum field (one with an infinite band limit) where the width of the averaging window corresponds to the size of the pixel.}

### 2018-07-16 articles/formalism/sections/real_space_formalism.tex
minor update to references, revision to discussion surround the radial kernel relation to zaldariagga kernel

We now define the different polarization data vectors and their representation in real and harmonic space as follows\footnote{We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables, as in \citep{Zaldarriaga2001a}. 
}:
%
\beqrys
\bar{S} &=& \bmat E \\ B  \emat_{2 \rm N_{\rm pix} \times 1};\qquad \bar{X} = \bmat _{+2}X \\ _{-2}X \emat_{2 \rm N_{\rm pix} \times 1};\qquad \bar{P} =\fqu_{\tiny {2 \rm N_{\rm pix} \times 1}} \,, \\
\tilde{S} &=& \bmat a^{E} \\ a^{B} \emat _{2 \rm N_{\rm alms} \times 1};\qquad \tilde{X} = \bmat _{+2} \tilde{X} \\ _{-2} \tilde{X} \emat_{2 \rm N_{\rm alms} \times 1} \,.
\eeqrys
%
The symbols have the same meaning as in \sec{sec:pol-primer}, except that the subscript ${\ell m}$ for the spherical harmonic coefficients is suppressed for cleaner notation.

@@

\section{Real space polarization operators} \label{sec:real_space_operators}
\subsection{Evaluating scalar $E/B$ from Stokes $Q/U$}\label{sec:qu2eb}
In \sec{sec:pol-primer} we described the conventional procedure of computing the scalar fields $E/B$ from the Stokes parameters $Q/U$. 
%To reiterate, this process involved taking the spin harmonic transform of the complex spin-2 fields ${}_{\pm2} \bar X$, forming specific linear combinations of the resultant coefficients of expansion ${}_{\pm 2} \tilde X_{\ell m}$ and evaluating the forward spin-0 transform to derive the scalar E \& B fields. 
In this section we derive the real space kernels which we can use to directly evaluate the scalar fields $E$ and $B$ on the sphere.  We use the vector-matrix notation introduced in \sec{sec:mat_pol_intro} to write down an operator equation relating the real space vector of scalars \vs to the Stokes polarization vector \vp{}:
%
\beqrys
\bar{S} &=& {{}_0\mathcal{Y}} \, \tilde T^{-1}  \, {{}_2\mathcal{Y}^{\ddagger}} \, \bar T  \bar{P}
= \frac{1}{2} {{}_0\mathcal{Y}} \, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \bar{P} \,,   \\
&=&  \bar O \bar{P}. \label{eq:qu2eb_op}
\eeqrys
%
The explicit form of the real space operator $\bar O$ can be derived by contracting over all the matrix operators. This procedure is explicitly worked out in the following set of equations:
%
\beqrys
\bar{O} &=& \frac{1}{2} {{}_0\mathcal{Y}}\, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \,, \\
&=& -0.5 \Delta \Omega \yzmat{e} \qutoxd \ymatc{q} \qutox   \,, \\
&=& -0.5 \Delta \Omega \begin{bmatrix} \sum ({}_{0}Y_e ~{}_{2}Y^{\dag}_q  +  {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & {\rm i}  \sum ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \\  - {\rm i} \sum  ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & \sum ({}_{0}Y_e~ {}_{2}Y^{\dag}_q + {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \end{bmatrix} \,, \label{eq:qu2eb_ker_1}
\eeqrys
%
where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{\dag}_q$ is used to denote the transposed conjugated matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. As before, we use the notation that the index $e$ denotes the location where the scalar fields are being evaluated, and the index $q$ denotes the location from which  the Stokes parameters are being accessed. Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following identity holds true:
%
\beq
 \left [\sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{+2}Y^*_{\ell m}(\hat{n}_q)\right]^* = \sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{-2}Y^*_{\ell m}(\hat{n}_q) \,,
 \eeq
 %
where the terms on either side of the equation are those that appear in \eq{eq:qu2eb_ker_1}. Note that the operator $\bar{O}$ is real as one expects, since each sub-matrix in \eq{eq:qu2eb_ker_1} is formed by summing a complex number and its conjugate.

>>>>

We now define the different polarization data vectors and their representation in real and harmonic space as follows\footnote{We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables. 
}:%as in \citep{Zaldarriaga2001a}
%
\beqrys
\bar{S} &=& \bmat E \\ B  \emat_{2 \rm N_{\rm pix} \times 1};\qquad \bar{X} = \bmat _{+2}X \\ _{-2}X \emat_{2 \rm N_{\rm pix} \times 1};\qquad \bar{P} =\fqu_{\tiny {2 \rm N_{\rm pix} \times 1}} \,, \\
\tilde{S} &=& \bmat a^{E} \\ a^{B} \emat _{2 \rm N_{\rm alms} \times 1};\qquad \tilde{X} = \bmat _{+2} \tilde{X} \\ _{-2} \tilde{X} \emat_{2 \rm N_{\rm alms} \times 1} \,.
\eeqrys
%
The symbols have the same meaning as in \sec{sec:pol-primer}, except that the subscript ${\ell m}$ for the spherical harmonic coefficients is suppressed for cleaner notation.

@@

\section{Real space polarization operators} \label{sec:real_space_operators}
\subsection{Evaluating scalars $E/B$ from Stokes $Q/U$}\label{sec:qu2eb}
In \sec{sec:pol-primer} we described the conventional procedure of computing the scalar fields $E/B$ from the Stokes parameters $Q/U$. 
%To reiterate, this process involved taking the spin harmonic transform of the complex spin-2 fields ${}_{\pm2} \bar X$, forming specific linear combinations of the resultant coefficients of expansion ${}_{\pm 2} \tilde X_{\ell m}$ and evaluating the forward spin-0 transform to derive the scalar E \& B fields. 
In this section we derive the real space kernels which we can use to directly evaluate the scalar fields $E$ and $B$ on the sphere.  We use the vector-matrix notation introduced in \sec{sec:mat_pol_intro} to write down an operator equation relating the real space vector of scalars \vs to the Stokes polarization vector \vp{}:
%
\beqrys
\bar{S} &=& {{}_0\mathcal{Y}} \, \tilde T^{-1}  \, {{}_2\mathcal{Y}^{\ddagger}} \, \bar T  \bar{P}
= \frac{1}{2} {{}_0\mathcal{Y}} \, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \bar{P} \,,   \\
&=&  \bar O \bar{P}. \label{eq:qu2eb_op}
\eeqrys
%
The explicit form of the real space operator $\bar O$ can be derived by contracting over all the matrix operators. This procedure is explicitly worked out in the following set of equations:
%
\beqrys
\bar{O} &=& \frac{1}{2} {{}_0\mathcal{Y}}\, \tilde T^{\dagger} {{}_2\mathcal{Y}^{\ddagger}} \, \bar T \,, \\
&=& -0.5 \Delta \Omega \yzmat{e} \qutoxd \ymatc{q} \qutox   \,, \\
&=& -0.5 \Delta \Omega \begin{bmatrix} \sum ({}_{0}Y_e ~{}_{2}Y^{\dag}_q  +  {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & {\rm i}  \sum ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \\  - {\rm i} \sum  ({}_{0}Y_e ~ {}_{2}Y^{\dag}_q - {}_{0}Y_e~ {}_{-2}Y^{\dag}_q) & \sum ({}_{0}Y_e~ {}_{2}Y^{\dag}_q + {}_{0}Y_e ~{}_{-2}Y^{\dag}_q)  \end{bmatrix} \,, \label{eq:qu2eb_ker_1}
\eeqrys
%
where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{\dag}_q$ is used to denote the transposed conjugated matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. As before, we use the notation that the index $e$ denotes the location where the scalar fields are being evaluated, and the index $q$ denotes the location from which  the Stokes parameters are being accessed. Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following identity holds true:
%
\beq
 \left [\sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{+2}Y^*_{\ell m}(\hat{n}_q)\right]^* = \sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{-2}Y^*_{\ell m}(\hat{n}_q) \,,
 \eeq
 %
where the terms on either side of the equation are those that appear in \eq{eq:qu2eb_ker_1}. Note that the operator $\bar{O}$ is real as one expects, since each sub-matrix in \eq{eq:qu2eb_ker_1} is formed by summing a complex number and its conjugate.
